Improved state-space models for inference about spatial and temporal variation in abundance from count data: Response to revision comments

### 2014-11-04 improved_state_space.tex
continuing to address new reviewer comments

The DM model is an extension of the 
\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance and detection parameters, 
the original $N$-mixture model uses replicate
observations at each site. The observations are collected during sufficiently
short time intervals such that the population can safely be
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

Like traditional state-space models, the DM model includes 
three conditionally related processes corresponding to: 
(1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted $N_{i,1}$, (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process 
describes the relationship between
abundance and the observed count data.

@@

The DM model assumes that abundance at time $t>1$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate, which can depend on $N_{i,t-1}$.
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new 
individuals; the ``autoregressive'' model, $G_{i,t} \sim
\mathrm{Pois}(\gamma N_{i,t-1})$
of geometric or density independent growth; and the
no-trend model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be accommodated, e.g., 
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
Partitioning population growth into survival and recruitment
is useful  
in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model is the simplest: 
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the 
Ricker model \citep{ricker:1954}. Another option is a 
modified Gompertz-logistic
density-dependent model \citep[henceforth Gompertz model,][]{hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.

@@

One limitation of the geometric-recruitment, exponential,
Ricker, and Gompertz versions of the DM models 
is that they include no mechanism for a local population to recover
after extinction. 
However, these models can be generalized 
to include both internal and external (immigration) contributions
to population growth. For example, an exponential plus immigration
model is:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
\end{linenomath*}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  
The geometric-recruitment, Ricker,
and Gompertz models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker + immigration and the 
Gompertz + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.

@@

\subsection*{Environmental and demographic stochasticity}
\label{subsec:stoch}
The original DM models allows for demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- 
via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
growth extensions use a single Poisson distribution for abundance. 
Conceptually, this is an appropriate method of modeling demographic stochasticity 
 for semelparous organisms with one generation per sampling interval (year)
 \citep{bonsall_hastings:2004},
and the DM models appropriate for modeling the demographic stochasticity of iteroparous organisms,
if they are underdispersed compared to the Poisson (but see Discussion).
 Dynamics of semelparous organisms with more than one generation per year 
 (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
 year is known, then this can be accounted for in our models by adding extra $N$'s between each
 count (see Supplement 1).

@@

We simulated data for 100 sites over 40 years, with 1000 simulations of 
each combination of parameter values.  All
simulations assumed initial abundance was Poisson distributed
and no covariates affected initial abundance, dynamics, or
detection probability.  Our first series of simulations
assumed dynamics were exponential.  We ran 
each combination of low, medium, and high $\Lambda \in
\{1,5,10\}$, $r \in \{-0.005, 0, 0.005\}$, and
$p \in \{0.05, 0.25, 0.5\}$. 
Our second series of simulations changed dynamics to the Ricker
model. We used an initial abundance of 10 and a detection probability
of 0.25, and simulated low, medium, and high values of equilibrium
abundance and maximum growth rate, (Table A1;
9 total combinations). Our third series of simulations was based
on the Ricker + immigration dynamics model; here we fixed all
parameters the same as the Ricker model (with $r$ = 0.05 and $K$ = 10) and
simulated low, medium, and high values of immigration rate (Table A1).

@@

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework with vague  
priors ($\Lambda$,
$\alpha$, and $K  \sim \mathrm{U}(0, 200)$, $r  \sim \mathrm{U}(0, 5)$,
$\iota  \sim \mathrm{U}(0, 15)$, and logit-linear $p$ coefficients $\sim \mathrm{N}(0, 100)$).  
We added random observer 
effects and environmental stochasticity, modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}.
We tested for lack of convergence using
five Markov chains for each model \citep{gelman_rubin:1992}.
For each chain, we drew at least 40,000 samples, 
after at least 3,000 adaptation iterations.   
We also used the random observer effects model to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we used a single chain.

@@

Estimates for the top ranked model 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved (Appendix B, Fig. B1). 
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically
and the estimate for r was intermediate. The 
estimate of the intercept for $p$ dropped. 
When environmental stochasticity was added, the estimate for $K$ increased. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects. 
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided little evidence of lack of convergence for any of the Bayesian models ran.

We estimated ovenbird abundance by route and year (and averaged over routes) 
using the Bayesian model with random observer effects (Fig.~\ref{fig:oven_N}).  
Estimated average abundance over time alternated between growing and stable periods.  
Estimated ovenbird
abundance was highest in western and eastern Maryland in 1970 and 1980, but became
more evenly distributed by 2010.

@@

One strategy for adhering to the original formulation of the
model that included survival and recruitment parameters would be to
use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
Bayesian analysis existing estimates of demographic parameters
could be used as informative priors, which might make it possible to
seperate the effects of movement versus births and deaths.
Alternatively, ecologists could combine count data with demographic data,
often referred to as an integrated population model
\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
We believe that a DM modeling framework could work well in this context,
and that the DM model could be viewed as a specific case of the
IPM proposed by \citet{buckland_etal:2004}.
However, we caution that when count data are affected by
movement and vital rates, it would be unwise to ignore the movement
processes.

Integrated population models may be viewed as the gold standard in
state-space modeling, but ecologists do not always have direct
information about vital rates, especially at large spatial scales.
Count data are much more common, relatively inexpensive to obtain, and
are produced by many of the largest monitoring programs in the
world, such as the BBS. 
In the absence of direct information about demographic parameters, and
when the original assumptions of the model do not hold, our extensions to the
DM model can still be used for important purposes such as predicting future
population size and changes in species distributions.

@@

Our emphasis was on increasing the practical utility of this
class of models, and so we avoided several conceptually interesting
extensions that we believe would be computationally prohibitive in many
cases. Nonetheless, we will discuss one --- spatially-explicit models of
immigration. 
In our models, immigration is currently modeled as independent of
population sizes in other sites. Several alternatives suggest
themselves if we think about movement in a metapopulation
context \citep{hanski:1998}: immigration conditional on total or mean
abundance across sites the previous time step; immigration
conditional on abundance in nearby sites only; and immigration
conditional on abundance at sites as a function of their distance
from the receiving site \citep{hastings:1991,hanski:1998}. These models would
require that all sites in the metapopulation are sampled, or that abundance at 
unsampled locations can be estimated \citep{lele_etal:1998}. 
Initial tests suggest that it would be
possible to fit some or all of these models, at least in the
Bayesian framework.

Our models generally estimated parameters well in analysis of data simulated with the same model.
Two exceptions were the estimates of instantaneous growth rate in exponential model when the 
true value of initial abundance was low, and the estimates of equilibrium abundance and maximum
growth rate in the Ricker model when the true value of maximum growth rate was low.  
The failure of the first case was likely due to a large number of sites starting at abundance 0 or reaching it early in the simulation.  
Because a site that hits abundance 0 in this simulation stayed at 0, this provided little information with
which to estimate growth rate.  The failure of the second case is also reasonable: with very
slow movement toward the equilibrium abundance comes little information for estimating
that abundance or the rate of growth.  In other cases, bias was low and coverage nominal.

@@

We thank J. Andrew Royle, T. Scott Sillett, and John Sauer for helpful suggestions on our analysis. 
We gratefully acknowledge the U. S. National Park 
Service, the Smithsonian Institution, and USGS's Breeding Bird Survey program for funding support.
We thank Keith Pardieck and David Ziolkowski for help with data acquisition. We are
grateful for the numerous helpful suggestions for manuscript
revision provided by J. Andrew Royle, T. Scott Sillett, and an anonymous
referee.

@@

\begin{figure}
\caption{}
  \centering
  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year6}
\label{fig:oven_N}
\end{figure}
\end{flushleft}
\end{spacing}
\end{document}

>>>>

The DM model is an extension of the 
\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance and detection parameters, 
the $N$-mixture model uses replicate % original
observations at each site. The observations are made during sufficiently
short time intervals such that the population can be %safely
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

Like traditional state-space models, the DM model includes 
three conditionally related processes corresponding to: 
(1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted $N_{i,1}$, (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process. % \citep{dail_madsen:2011}.
The first two processes describe the %the state process---
variation in abundance in space and time. The third process 
describes the relationship between
abundance and the observed count data.

@@

The DM model assumes that abundance at time $t>1$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate, which can depend on $N_{i,t-1}$.
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$; %and which simulates a ``propagule rain'' of new individuals; 
the ``autoregressive'' model, $G_{i,t} \sim
\mathrm{Pois}(\gamma N_{i,t-1})$
of geometric or density independent growth; and the
no-trend model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be accommodated, e.g., 
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
Partitioning population growth into survival and recruitment
is useful  
in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to movement.
%immigration and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model, which describes population growth in an 
unlimited environment, is the simplest: 
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic or maximum per capita
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the 
Ricker model \citep{ricker:1954}. In the Ricker model, 
density dependence becomes stronger as abundance increases well above $K$,
leading to overcompensation    
%due to the exponential function relating to carrying capacity 
\citep[abundance goes from above $K$ to below it;][]{sabo_etal:2004}.
Another option is a 
modified Gompertz-logistic
density-
dependent model \citep[henceforth Gompertz model;][]{gompertz:1825,
hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker model, but dynamics can only be overcompensatory at high values of $r$
(approximately $r > log(K)$).
%population growth declines linearly with an increase in the natural logarithm of abundance,
%rather than linearly with an increase in abundance. 
Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.

@@

One limitation of the geometric-recruitment, exponential,
Ricker, and Gompertz versions of the DM models 
is that they include no mechanism for a local population to recover
after extinction. 
However, these models can be generalized 
to allow immigration to contribute
to population growth. For example, an exponential plus immigration
model is:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
\end{linenomath*}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  
The geometric-recruitment, Ricker,
and Gompertz models can be extended to allow for immigration in the
same way.  Because the population contribution from immigration does
not depend on current abundance, $\iota$ and other growth parameters
are separately estimable.
 Deterministic versions of the Ricker + immigration and the 
Gompertz + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.

@@

\subsection*{Environmental and demographic stochasticity}
\label{subsec:stoch}
The original DM models allows for demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- 
via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
growth extensions use a single Poisson distribution for abundance. 
Conceptually, this is an appropriate method of modeling demographic stochasticity 
 for semelparous populations with one generation per sampling interval (year)
 \citep{bonsall_hastings:2004},
and the DM models appropriate for modeling the demographic stochasticity of iteroparous populations,
if they are underdispersed compared to the Poisson (but see Discussion).
 Dynamics of semelparous populations with more than one generation per year 
 (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
 year is known, then this can be accounted for in our models by adding extra $N$'s between each
 count (see Supplement 1).

@@

We simulated data for 100 sites over 40 years, with 1000 simulations of 
each combination of parameter values.  All
simulations assumed initial abundance was Poisson distributed
and no covariates affected initial abundance, dynamics, or
detection probability.  Our first series of simulations
assumed dynamics were exponential (Eq.~\ref{eq:exp}).  We ran 
each combination of low, medium, and high initial abundance, 
growth rate, and detection probability ($\Lambda \in
\{1,5,10\}$, $r \in \{-0.005, 0, 0.005\}$, and
 $p \in \{0.05, 0.25, 0.5\}$). 
Our second series of simulations changed dynamics to the Ricker
model (Eq.~\ref{eq:rick}). We used $\Lambda$ of 10 and $p$
of 0.25, and simulated low, medium, and high values of equilibrium
abundance $(K)$ and $r$ (Table A1;
9 total combinations). Our third series of simulations was based
on the Ricker + immigration dynamics model; here we fixed all
parameters the same as the Ricker model (with $r$ = 0.05 and $K$ = 10) and
simulated low, medium, and high values of immigration rate ($\iota$) (Table A1).

@@

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework with vague  
priors ($\Lambda$,
$\alpha$, and $K  \sim \mathrm{U}(0, 200)$, $r  \sim \mathrm{U}(0, 5)$,
$\iota  \sim \mathrm{U}(0, 15)$, and logit-linear $p$ coefficients $\sim \mathrm{N}(0, 100)$).  
We added random observer 
effects and environmental stochasticity, modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}.
We also used the random observer effects model to obtain route and 
year specific abundance estimates.    
We tested for lack of convergence using four or
five Markov chains for each model \citep{gelman_rubin:1992}.
For each chain, we drew at least 40,000 samples, 
after at least 3,000 adaptation iterations.

@@

Estimates for the top ranked model 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved (Appendix B, Fig. B1). 
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically
and the estimate for r was intermediate. The 
estimate of the intercept for $p$ dropped. 
When environmental stochasticity was added, the estimate for $K$ increased. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects.

We estimated ovenbird abundance by route and year (and averaged over routes) 
using the Bayesian model with random observer effects (Fig.~\ref{fig:oven_N}).  
Estimated average abundance over time alternated between growing and stable periods.  
Estimated ovenbird
abundance was highest in western and eastern Maryland in 1970 and 1980, but became
more evenly distributed by 2010.  Gelman and Rubin diagnostics and visual 
examination of the chain trajectory and density plots
provided little evidence of lack of convergence for any of the Bayesian models ran.

@@

%One strategy for adhering to the original formulation of the
%model that included survival and recruitment parameters would be to
%use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
%Bayesian analysis existing estimates of demographic parameters
%could be used as informative priors, which might make it possible to
%seperate the effects of movement versus births and deaths.
%Alternatively, ecologists could combine count data with demographic data,
%often referred to as an integrated population model
%\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
%We believe that a DM modeling framework could work well in this context,
%and that the DM model could be viewed as a specific case of the
%IPM proposed by \citet{buckland_etal:2004}.
%However, we caution that when count data are affected by
%movement and vital rates, it would be unwise to ignore the movement
%processes.
%
%Integrated population models may be viewed as the gold standard in
%state-space modeling, but ecologists do not always have direct
%information about vital rates, especially at large spatial scales.
%Count data are much more common, relatively inexpensive to obtain, and
%are produced by many of the largest monitoring programs in the
%world, such as the BBS. 
%In the absence of direct information about demographic parameters, and
%when the original assumptions of the model do not hold, our extensions to the
%DM model can still be used for important purposes such as predicting future
%population size and changes in species distributions.

@@

Our emphasis was on increasing the practical utility of this
class of models, and so we avoided several conceptually interesting
extensions that we believe would be computationally prohibitive in many
cases. Nonetheless, we will discuss one --- spatially-explicit models of
immigration. 
In our models, immigration is currently modeled as independent of
population sizes in other sites, and emigration is included only
as an implicit piece of apparent survival or population growth. 
We suggest alternatives that put movement in a metapopulation
context: immigration conditional on total or mean % \citep{hanski:1998}
abundance across sites the previous time step; immigration
conditional on abundance in nearby sites only; and immigration
conditional on abundance at sites as a function of their distance
from the receiving site \citep{hastings:1991,hanski:1998}. These models would
require that all sites in the metapopulation are sampled, or that abundance at 
unsampled locations can be estimated \citep{lele_etal:1998}. 
Initial tests suggest that it would be
possible to fit some or all of these models, at least in the
Bayesian framework. It may also be possible to model an explicit emigration
process or rate and tie immigration at nearby sites to this rate.

Our models generally estimated parameters well in analysis of data simulated with the same model.
Two exceptions were the estimates of instantaneous growth rate in exponential model when the 
true value of initial abundance was low, and the estimates of equilibrium abundance and maximum
growth rate in the Ricker model when the true value of maximum growth rate was low.  
The failure of the first case was likely due to a large number of sites starting at abundance 0 or reaching it early in the simulation.  
Because a site that hits abundance 0 in this simulation stayed at 0, this provided little information with
which to estimate growth rate.  The failure of the second case is also reasonable: with very
slow movement toward the equilibrium abundance comes little information for estimating
that abundance or the rate of growth.  In other cases, bias was low and coverage nominal.
However, we simulated 40 years of data; with fewer years of data these models might
experience more problems.

@@

We thank J. Andrew Royle, T. Scott Sillett, and John Sauer for helpful suggestions on our analysis. 
We gratefully acknowledge the U. S. National Park 
Service, the Smithsonian Institution, and USGS's Breeding Bird Survey program for funding support.
We thank Keith Pardieck and David Ziolkowski for help with data acquisition. We are
grateful for the numerous helpful suggestions for manuscript
revision provided by J. Andrew Royle, T. Scott Sillett, and anonymous referees.

@@

\begin{figure}
\caption{}
  \centering
  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year8}
\label{fig:oven_N}
\end{figure}
\end{flushleft}
\end{spacing}
\end{document}

### 2014-11-04 response to revision comments.tex
continuing to address new reviewer comments

\vspace{0.5cm}
\textit{[I can do this.]}
\vspace{0.5cm}

@@

I find that basing the introduction upon the notion of state-based model is rather confusing
regarding the real goal of the paper (see above). The Dail and Madsen model and statespace
models are defined in two different ``statistical worlds''. On the one hand, the constant
Dail and Madsen model (2011) has a Markov property so the variables can only take on a
finite set of discrete values for a finite period of observation. The distinguish feature of
Markov chains is that the conditional probability of $x_{t}$ given the entire past depends only on
the immediately preceding variable $x_{t-1}$. In addition, the errors of observation are well
identified because they are related to imperfect detection of individuals and expressed with
a discrete probability distribution. On the other hand, the state-space model framework
provides a structure for extending time-series models to handle both observation and
process errors. The term ``state-space'' model refers to models where the process model has
an autoregressive property and where the errors have a Gaussian distribution.
Autoregressive model are defined for continuous variable, and so the errors are normally
distributed. A space-state model, being given its properties, can be estimated by using a
Kalman filter. The Kalman filter is an updating procedure whereby the prior estimate of the
state is corrected by how well the next observation can be predicted.

@@

\vspace{0.5cm}
\textit{[I can work on this.]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[I can see the reviewer's point (there's no actual mainland), although I think it really
behaves the same way.  Should I just cut/change this?]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[I can do this.]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[I can work on this.]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[I can work on these two.]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[I can work on this.  Should I just speculate in the discussion about fewer years
of data?  I would think less precision would be the result.  Fewer than 10 years of data
could especially be a problem for density-dependent models, but again that's speculation.]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[I can work on this.  Not sure yet what we can/should cut though.]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[Between this comment and the one by the other reviewer, sounds like we should 
put all three of these figures in the main text.  If we still have to keep it to 30 pages,
that requires at least two pages of cutting, which is a lot.  Probably more, since we'll also 
be adding a few things.  This reviewer suggests cutting
text from DM model description and discussion.  I will work on those, but maybe we
should also consider emailing Evan Cooch about the length limit.]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[I can work on this.  Not sure yet what the answer is though.]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[Will this work?  I should try it.]}
\vspace{0.5cm}

>>>>

\vspace{0.5cm}
\textit{[We have described the classical population growth models in more 
detail (lines ).]}
\vspace{0.5cm}

@@

I find that basing the introduction upon the notion of state-based model is rather confusing
regarding the real goal of the paper (see above). The Dail and Madsen model and state-space
models are defined in two different ``statistical worlds''. On the one hand, the constant
Dail and Madsen model (2011) has a Markov property so the variables can only take on a
finite set of discrete values for a finite period of observation. The distinguish feature of
Markov chains is that the conditional probability of $x_{t}$ given the entire past depends only on
the immediately preceding variable $x_{t-1}$. In addition, the errors of observation are well
identified because they are related to imperfect detection of individuals and expressed with
a discrete probability distribution. On the other hand, the state-space model framework
provides a structure for extending time-series models to handle both observation and
process errors. The term ``state-space'' model refers to models where the process model has
an autoregressive property and where the errors have a Gaussian distribution.
Autoregressive model are defined for continuous variable, and so the errors are normally
distributed. A space-state model, being given its properties, can be estimated by using a
Kalman filter. The Kalman filter is an updating procedure whereby the prior estimate of the
state is corrected by how well the next observation can be predicted.

@@

\vspace{0.5cm}
\textit{[Richard, any thoughts on this?  I can see the reviewer's point - there are a lot of DM
papers out now.  And this is almost 3 pages.  But I like what we've got here, and if we cut or
move it we'd have to be careful, because we refer to this section a good deal in our extensions.
How we shorten it may depend on how we respond to the DM model as state-space model critique.
Please take a stab at this if you have time and ideas on how to shorten or move.]}
\vspace{0.5cm}

@@

@@

\vspace{0.5cm}
\textit{[We have described the Gompertz model in more 
detail (lines ).]}

\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{We have edited the text for clarity (lines ).  We have added discussion about emigration
(lines ).}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{We were referring to generations of the population or species, not the individual.  We have edited 
this paragraph for clarity.}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{We have added clarification and equation references to this paragraph. We expect that
the models would not perform as well with far fewer years of data, and have commented on
this in the discussion (lines).}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[Richard, I commented out the two paragraphs on IPMs.  What do you think?]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{[Between this comment and the one by the other reviewer, sounds like we should 
put all three of these figures in the main text.  If we still have to keep it to 30 pages,
that requires a page and a half of additional cutting, which is a lot.  This reviewer suggests cutting
text from DM model description and discussion.  I have worked on those (see above), but maybe we
should also consider emailing Evan Cooch about the length limit.]}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{We believe the reviewer has hit upon it - the population contribution from immigration
does not depend on current abundance, whereas the contributions from $r$ in the density-dependent
models and from $\gamma$ in the geometric-recruitment model do (in different ways).  With sufficient
temporal and spatial replication, these parameters are separately estimable.  We have added text to that
effect (lines ).}
\vspace{0.5cm}

@@

\vspace{0.5cm}
\textit{We now have multiple chains for this model as well, and the convergence is good
with very similar results.}
\vspace{0.5cm}
